---
title: "MyRWA Database Functions"
author: "Jeffrey D. Walker, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
base_dir <- 'C:/Users/Jeff/Dropbox/Hopper'
```

The `myrwaR` package includes a set of functions for retrieving data from the MyRWA Water Quality Access Database. To load the package:

```{r load-package}
library(myrwaR)
```

## Connecting to the Database

To connect to the database, use `db_connect()` and specify the path to the database, which returns a connection handler that is later used to retrieve the data.

```{r connect}
ch <- db_connect(path = file.path(base_dir, "MysticDB_20160204.accdb"))
```

## Retrieving Data from Any Table

The `db_list_tables()` function lists the names of tables in the database (note that `dplyr` has 
a function with the same name, so we need to namespace it with `myrwaR::`):

```{r list-tables}
myrwaR::db_list_tables(ch)
```

To load the data from any one of these tables, use `db_table()`. For example, to get the `Location` table:

```{r locations}
locations <- db_table(ch, table_name = "Location")
str(locations)
```


## Retrieving Water Quality Data

If a database connection has already been created, water quality data can be retrieved using `db_results()`. This function will automatically join the `Visit` and `Result` tables to provide the location IDs, datetimes, and useful attributes for each row in the `Result` table.

```{r results}
results <- db_results(ch)
```

## Water Quality Convenience Function

Finally, to connect to the database and retrieve the water quality data in one fell swoop, use the `load_wq()` function. The function will connect to the database, and then merge the `Result`, `Visit`, and `Location` database tables into a single dataframe.

```{r load-wq}
wq <- load_wq(path = file.path(base_dir, "MysticDB_20160204.accdb"), sample_types = c("S"), exclude_flags = TRUE)
str(wq)
```

The `sample_types` parameter lets you specify which sample types to include in the data frame. By default, it will only return routine samples ("S"). To include other sample types such as field duplicates and field blanks, just add other SampleType IDs `sample_types = c("S", "FD", "FB")`. To see a complete list of sample types, just load the `SampleType` table: `db_table(ch, table_name="SampleType")`.

The `exclude_flags` parameter specifies whether the resulting data frame should exclude flagged samples (default is `TRUE`). To include the flagged samples, set this to `FALSE`.

## Closing the Connection

To close a connection, simply use the `close()` function:

```{r close}
close(ch)
```

# QAQC Procedures

## Version Comparison

The `compare_database()` function summarizes changes between to database versions. This function currently loads the Result, Visit, and Location tables and identifies any rows that have been added or removed. Note that this does not look at changes to values of existing rows, only row additions and removals.

```{r compare}
compare_database(old_path = file.path(base_dir, "MysticDB_20160120.accdb"),
                 new_path = file.path(base_dir, "MysticDB_20160204.accdb"))
```

The `compare_database()` function can also save the output to a text file by setting the `log_file` argument.

```{r compare-log, eval=FALSE}
compare_database(old_path = file.path(base_dir, "MysticDB_20150227.accdb"),
                 new_path = file.path(base_dir, "MysticDB_20150529.accdb"),
                 log_file = file.path(base_dir, 'compare_20150227_20150529.txt'))
```
