---
title: "MyRWA Database Functions"
author: "Jeffrey D. Walker, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Database Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
```

The `myrwaR` package includes a set of functions for retrieving data from the MyRWA Water Quality Access Database. First, we need to load the package:

```{r load-package}
library(myrwaR)
```

# Accessing the Database

## Connecting to the Database

To connect to the database, use `db_connect()` and specify the path to the database, which returns a connection handler that is later used to retrieve the data.

```{r connect}
base_dir <- 'D:/Dropbox/Work/mystic/db'
db_path <- file.path(base_dir, "MysticDB_20160208.accdb")
ch <- db_connect(path = db_path)
```

## Retrieving Data from Any Table

The `db_list_tables()` function lists the names of tables in the database (note that `dplyr` has 
a function with the same name, so you may need to namespace it with `myrwaR::`):

```{r list-tables}
myrwaR::db_list_tables(ch)
```

To load the data from any one of these tables, use `db_table()`. For example, to get the `Location` table:

```{r locations}
locations <- db_table(ch, table_name = "Location")
str(locations)
```


## Retrieving Water Quality Data

If a database connection has already been created, water quality data can be retrieved using `db_results()`. This function will automatically join the `Visit` and `Result` tables to provide the location IDs, datetimes, and useful attributes for each row in the `Result` table. Note that this function will not join the `Location` table, and thus not contain latitudes and longitudes. See the `load_wq()` function in the next section.

```{r results}
results <- db_results(ch)
```

## Water Quality Dataset Function

Finally, to connect to the database and retrieve the water quality data with the location information, use the `load_wq()` function (note that the name of this function differs from other database functions to maintain consistency with older R scripts used by MyRWA). The `load_wq()` function will connect to the database, and then merge the `Result`, `Visit`, and `Location` database tables into a single dataframe.

```{r load-wq}
wq <- load_wq(path = db_path, sample_types = c("S"), exclude_flags = TRUE)
str(wq)
```

The `sample_types` parameter lets you specify which sample types to include in the data frame. By default, it will only return routine samples (`"S"`). To include other sample types such as field duplicates and field blanks, just add other SampleType IDs `sample_types = c("S", "FD", "FB")`, or set `sample_types = NULL` to retrieve all sample types. To see a complete list of sample types, load the `SampleType` table: `db_table(ch, table_name="SampleType")`.

The `exclude_flags` parameter specifies whether the resulting data frame should exclude flagged samples (default is `TRUE`). To include the flagged samples, set this to `FALSE`.

## Closing the Connection

To close a connection, simply use the `close()` function:

```{r close}
close(ch)
```

# QAQC Procedures

## Version Comparison

The `compare_database()` function summarizes changes between to database versions. This function loads the `Result`, `Visit`, and `Location` tables from each database version, and then summarizes how many rows have been added or removed from the first to the second version. Note that this does not look at changes to the values of each row, it only checks for row additions and removals.

```{r compare}
compare_database(old_path = file.path(base_dir, "MysticDB_20160120.accdb"),
                 new_path = file.path(base_dir, "MysticDB_20160208.accdb"))
```

The output of the `compare_database()` function can also be saves to a text file by setting the `log_file` argument.

```{r compare-log, eval=FALSE}
compare_database(old_path = file.path(base_dir, "MysticDB_20150227.accdb"),
                 new_path = file.path(base_dir, "MysticDB_20150529.accdb"),
                 log_file = file.path(base_dir, 'compare_20150227_20150529.txt'))
```

## Database QAQC Checks

The `db_qaqc_suite()` function runs a series of checks on the `Result` and `Visit` tables These checks are designed to find invalid values in each column of these tables. The types of checks include:

- inconsitencies in lower/upper case
- presence of empty strings, which should be replaced with NA's (e.g. `""`)
- presence of untrimmed strings that contain an extra space before or after the value (e.g. `"FD "`)
- presence of `NA` in columns that should not have missing values 
- invalid date ranges (all dates should be between 1900-01-01 and the today's date)
- invalid relationships between tables (e.g. `Result.Units->Units.ID`, `Result.FlagID->ResultFlag.ID`, `Visit.SampleTypeID->SampleType.ID`, and `Visit.ProjectID->Project.ID`)

Note that the first three checks are included because Microsoft Access does not enforce consistency across these issues and thus may go unnoticed within the database. However, these issues can cause problems when importing the results into R, which does require these consistencies.

For each set of checks, `db_qaqc_suite()` function will report which columns failed, the reason, and the number of rows that failed each check. The specific rows that failed can be shown by setting `print.values=TRUE` (not shown here for brevity). 

```{r qaqc-suite}
ch <- db_connect(db_path)
db_qaqc_suite(ch, print.rows=FALSE)
close(ch)
```
