---
title: "Functions for Retrieving MyRWA Data from STORET"
author: "Jeffrey D. Walker, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STORET Retrieval Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
```

This document describes functions in the `myrwaR` package for handling STORET data. First, we'll walk through how to fetch MyRWA data that is currently in STORET. Then we will compare the STORET dataset to what is currently in the Access database.

```{r libraries, message = FALSE, warning = FALSE}
library(myrwaR)
library(lubridate)
library(dplyr)
library(tidyr)
```

# Fetching Data from STORET

## Locations Table

The MyRWA locations currently in STORET can be fetched using the `get_storet_locations()` function. This function will download a zip file from STORET, unzip the contents, and load the resulting `station.csv` into a data frame. Note that `verbose` is optional and `TRUE` by default, but you can set it to `FALSE` to suppress the log output.

```{r loc}
storet_locations <- get_storet_locations(verbose = TRUE)
str(storet_locations)
```

## Results Table

The MyRWA results table currently in STORET can be fetched using the `get_storet_results()` function. This function will download a zip file from STORET, unzip the contents, and load the resulting `results.csv` into a data frame. Note that the `verbose` argument can be used here too.

```{r res}
storet_results <- get_storet_results()
str(storet_results)
```

# Comparing STORET Data to MyRWA Database

This section compares the number of samples by different variables (e.g. location, characteristic, year) between STORET and the current water quality database.

First, we need to extract the dataset from the current database using the function `wqx_results()` (this will convert the dataset to WQX format, which is similar, though not identical, to the format provided by STORET):

```{r db-results}
ch <- db_connect("D:/Dropbox/Work/mystic/db/MysticDB_20160208.accdb")
db_results <- wqx_results(ch, projects='BASE')
close(ch)
```

Next, we'll convert the `ActivityStartDate` column in the `db_results` data frame to `Date` types so we can extract the Year later, and also prepend `"MYRWA-"` to the `MonitoringLocationID` and `ActivityID` columns since the data returned from STORET includes this prefix but the WQX export format does not.

```{r db-filter-dates}
db_results <- mutate(db_results,
                     ActivityStartDate    = as.Date(ActivityStartDate),
                     MonitoringLocationID = paste("MYRWA", MonitoringLocationID, sep = "-"),
                     ActivityID           = paste("MYRWA", ActivityID, sep = "-"))
```

## Compare Sample Counts

To compare the STORET dataset to that in the current database, we can use the `storet_compare_count()` function. This function takes two arguments, one for each dataset. Each argument should be a named list containing one or more columns from the original results table. The names in each list let you rename the columns from the original dataset (e.g. `LocationID = storet_results$MonitoringLocationIdentifier)`) and also define names for new columns (e.g. `Year = year(storet_results$ActivityStartDate)`). The names in the two list arguments (`storet` and `db`) must be identical and cannot be `"Source"` or `"n"`, which are created internally within the function. For each list, the function will count the number of samples for each unique combination of values, and then compares those counts between the two datasets to identify any differences.

### By Location ID

First, let's compare the number of samples by Location ID. Note how the names of the original columns are different in `storet_results` and `db_results`, and thus are renamed to both be `LocationID` within the list arguments.

```{r cnt-loc}
storet_compare_count(storet = list(LocationID = storet_results$MonitoringLocationIdentifier),
                     db     = list(LocationID = db_results$MonitoringLocationID))
```

### By Characteristic Name

Next, let's compare the number of samples by `CharacteristicName`:

```{r cnt-char}
storet_compare_count(storet = list(CharacteristicName = storet_results$CharacteristicName),
                     db     = list(CharacteristicName = db_results$CharacteristicName))
```

### By Year

And then, we'll compare counts by `Year`. Note how the columns are converted from dates to years in each list using the `year()` function. 

```{r cnt-yr}
storet_compare_count(storet = list(Year = year(storet_results$ActivityStartDate)),
                     db     = list(Year = year(db_results$ActivityStartDate)))
```

### By Date

Finally, we'll compare counts by `ActivityStartDate`:

```{r cnt-date}
storet_compare_count(storet = list(Date = storet_results$ActivityStartDate),
                     db     = list(Date = db_results$ActivityStartDate))
```

### Summary

These comparisons show that there are a number of sample events from 2015 that are in the database but not yet in STORET. This approach can thus be useful for figuring out which sample events should be exported for the next STORET upload. It's also important to recognize that the differences shown in these individual tables do not reflect individual discrepancies, but rather that there are a handful of sampling events that have not yet been uploaded to STORET. In other words, the differences in sample counts for individual locations or characteristics simply reflect that some of the visits have not been uploaded yet.

## Compare Sample Counts among STORET Visits Only

To verify that the counts by location and characteristic match among only the visits that have been uploaded to STORET, we can simply filter the database dataset to only include sample dates that are also in the STORET dataset.

```{r filter-db}
db_results <- filter(db_results, ActivityStartDate %in% unique(storet_results$ActivityStartDate))

storet_compare_count(storet = list(LocationID = storet_results$MonitoringLocationIdentifier),
                     db     = list(LocationID = db_results$MonitoringLocationID))
storet_compare_count(storet = list(CharacteristicName = storet_results$CharacteristicName),
                     db     = list(CharacteristicName = db_results$CharacteristicName))
storet_compare_count(storet = list(Year = year(storet_results$ActivityStartDate)),
                     db     = list(Year = year(db_results$ActivityStartDate)))
storet_compare_count(storet = list(Date = storet_results$ActivityStartDate),
                     db     = list(Date = db_results$ActivityStartDate))
```

Based on these comparisons using only the dates that are in STORET, we see one issue in the counts by location ID. There were 8 samples where the LocationID appears to have been changed in the database from BEI001 to BEI093. We can then identify the specific sample date associated with this difference by doing a comparison on both `Date` and `LocationID`.

```{r cnt-date-loc}
storet_compare_count(storet = list(Date       = storet_results$ActivityStartDate,
                                   LocationID = storet_results$MonitoringLocationIdentifier),
                     db     = list(Date       = db_results$ActivityStartDate,
                                   LocationID = db_results$MonitoringLocationID))
```

And we see that the difference in `LocationID` occurs in the `2015-01-23` sample visit. 

Another thing we can do is look for differences in the `ActivityID`, which shows the same result. Note that the lab (`L`) and field (`F`) samples each have different `ActivityIDs`, so there are four rows total, two for each dataset.

```{r cnt-activity-id}
storet_compare_count(storet = list(ActivityID = storet_results$ActivityIdentifier),
                     db     = list(ActivityID = db_results$ActivityID))
```

We now have some useful information to identify and correct the issue. Here are the corresponding rows from the STORET and Database result tables.

```{r filter-differences}
filter(storet_results,
       MonitoringLocationIdentifier == "MYRWA-BEI093",
       ActivityStartDate == as.Date("2015-01-23")) %>%
  select(ActivityIdentifier, ActivityStartDate, ActivityStartTime.Time,
         MonitoringLocationIdentifier, CharacteristicName, ResultMeasureValue,
         ResultMeasure.MeasureUnitCode)

filter(db_results,
       MonitoringLocationID == "MYRWA-BEI001",
       ActivityStartDate == as.Date("2015-01-23")) %>%
  select(ActivityID, ActivityStartDate, ActivityStartTime, MonitoringLocationID,
         CharacteristicName, ResultValue, ResultUnit)
```

